---
title: "MS4217: Lab 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
library(latex2exp)
library(expm)

write_matrix <- function(x) {
  begin <- "\\begin{pmatrix}"
  end <- "\\end{pmatrix}"
  X <-
    apply(x, 1, function(x) {
      paste(
        paste(x, collapse = "&"),
        "\\\\"
      )
    })
  paste(c(begin, X, end), collapse = "")
}

# P_index <- read.delim(file = './Fire_weather_index.prn', sep = ' ')
# P_wet <- read.delim(file = './Wetland_distribution.prn', sep = '|')

library(cowplot)
theme_set(theme_cowplot())

```

## Metastatic Cancer Progression 

The metastatic progression of lung cancer throughout the body is modeled in Newton et al. (2012). The 50 possible metastatic locations for cancer spread are the state space for a Markov chain model. Matrix entries were estimated from autopsy data extracted from 3,827 patients. The progress of the disease is observed as a random walk on the weighted, directed graph in Figure 2.6. Site 23 represents the lung. An important quantity associated with this model is the mean first-passage time, the average number of steps it takes to move from the lung to each of the other locations in the body.

Medical researchers can use simulation to study the progression of lung cancer in the body.

### Excerise 1:

a. Read in the data and examine the matrix. 
a. Examine its class. 
a. If needed convert the object you have into a matrix.
a. Formate the rows and columns of the matrix.

```{r read-in-data,echo=FALSE}
# your code goes here
P_lung_df <- read_csv(file = './lungcancer.csv') # read in the data
class(P_lung_df) # probably better to have this as a class! 
P_lung <- as.matrix(P_lung_df)

colnames(P_lung) # what are the states
rownames(P_lung) <- colnames(P_lung) # save the row names
```

### Excerise 2:

Extract the site names for each state in the matrix and generate a summary table using `kable` from the `knitr` package. 

```{r state-mapping}
state_names <- colnames(P_lung) # for reference create a vector of state_names (the labels)
state_names_df <- tibble(state_no = 1:length(state_names), state_names = state_names) # create a table
state_names_df %>% 
  knitr::kable(
    col.names = c('i','State $S_i$'), caption = 'State labels for cancer dataset.'
    ) # print the table to screen
```

### Excerise 3:

Create a function to simulate the path of a Markov chain.

```{r markov-function}
markov <- function(init, mat, n, labels){
  # one line if statement to add labels if not present
  if(missing(labels)) labels <- 1:length(init)
  
  simlist <- numeric(n+1) # where we are going to store each transition
  states <- 1:length(init) # How many states do we have?  
  simlist[1] <- sample(states,1,prob=init) # what state do we start in from the inital state?
  
  for(i in 2:(n+1)){ # from state i-1 where do we go next?
    # sample the prob vec for trans from x_{t-1} to X_{t}
    simlist[i] <- sample(states,1,prob=mat[simlist[i-1],]) 
  }
  return(labels[simlist])
}

```


### Excerise 4: 

a. Use the previous function to estimate the probability distribution for $X_8$. 
a. Visulate the probaility using a bar plot for each site. 
a. What are the three most common sites to observe cancer after 8 transitions? 

```{r}

init <- c(rep(0,22),1,rep(0,27))
n <- 8 

M = 1000
mc_results <- matrix(NA, nrow = M, ncol = n+1)
for(m in 1:M){
  mc_results[m,] <- markov(init, P_lung, n)
}

mc_X_10_df <- tibble(sim_no = 1:M, state_no = mc_results[,9]) %>% 
  left_join(y = state_names_df, by = 'state_no')

X_10_dist <- mc_X_10_df %>% count(state_names) %>% 
  mutate(prob = n/sum(n)) %>% 
  arrange(desc(prob))

mc_X_10_df %>% ggplot(aes(y = state_names)) + 
  geom_bar()

```



### Excerise 5:

a. Compare this to the exact probability using direct matrix multiplication. 
a. Comment of the difference between the simulation and exact probability. 

```{r}
library(expm)

P_10 <- P_lung %^% 5
exact <- tibble(state_names = names(P_10[23,]), exact_prob = P_10[23,])

X_10_dist <- left_join(X_10_dist, exact, by = 'state_names') %>% 
  mutate(diff = prob - exact_prob)

X_10_dist %>% ggplot(aes(x = diff)) + geom_density(fill = 'gold', alpha = 0.5) + 
  geom_vline(xintercept = 0, linetype = 'dashed', color = "red")

```


### Excerise 6:

a. Extend this above to 1000 MC draws from the Markov chain simulation to calculate the mean first passage. 
a. Generate a summary table of the results displaying the states with the 5 shortest mean first passage times.
a. What insight might you be able to give medical research. 

```{r}

M = 1000
n = 1000
mc_results <- matrix(NA, nrow = M, ncol = n+1)

for(m in 1:M){
  mc_results[m,] <- markov(init, P_lung, n)
}

# for each state what was the first time we came across it? 
first_passage <- matrix(NA, ncol = 50, nrow = M)
colnames(first_passage) <- state_names

for(m in 1:nrow(mc_results)){ # m <- 1
  for(s in 1:ncol(first_passage)){ # s <- 1
    first_passage[m,s] <- which(mc_results[m,] == s)[1] - 1
  }
}

first_passage[is.na(first_passage)] <- Inf


mfp <- apply(first_passage, 2, mean)
mfp_df <- tibble(state_names = state_names, mfp = mfp) %>% 
  arrange(mfp)

mfp_df %>% slice(1:5) %>% knitr::kable()

mfp_df %>% filter(is.finite(mfp)) %>% ggplot(aes(y = state_names, x = mfp)) + 
  geom_bar(stat="identity") + 
  xlab("Mean first passage time") +
  ylab(TeX('S_i'))
```


