---
title: "MS4217: Lab 3"
author: "Name & ID Number"
date: "Date of lab session"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
library(latex2exp)
library(expm)

write_matrix <- function(x) {
  begin <- "\\begin{pmatrix}"
  end <- "\\end{pmatrix}"
  X <-
    apply(x, 1, function(x) {
      paste(
        paste(x, collapse = "&"),
        "\\\\"
      )
    })
  paste(c(begin, X, end), collapse = "")
}

# P_index <- read.delim(file = './Fire_weather_index.prn', sep = ' ')
# P_wet <- read.delim(file = './Wetland_distribution.prn', sep = '|')

library(cowplot)
theme_set(theme_cowplot())

```

## Metastatic Cancer Progression 

The metastatic progression of lung cancer throughout the body is modeled in Newton et al. (2012). The 50 possible metastatic locations for cancer spread are the state space for a Markov chain model. Matrix entries were estimated from autopsy data extracted from 3,827 patients. The progress of the disease is observed as a random walk on the weighted, directed graph in Figure 2.6. Site 23 represents the lung. An important quantity associated with this model is the mean first-passage time, the average number of steps it takes to move from the lung to each of the other locations in the body.

Medical researchers can use simulation to study the progression of lung cancer in the body.

### Excerise 1:

a. Read in the data and examine the matrix. 
a. Examine its class. 
a. If needed convert the object you have into a matrix.
a. Formate the rows and columns of the matrix.

```{r read-in-data,echo=FALSE}
# code goes here


```

### Excerise 2:

Extract the site names for each state in the matrix and generate a summary table using `kable` from the `knitr` package. 

```{r state-mapping}
# code goes here

```

### Excerise 3:

Create a function to simulate the path of a Markov chain.

```{r markov-function}
# code goes here


```


### Excerise 4: 

a. Use the previous function to estimate the probability distribution for $X_8$. 
a. Visulate the probaility using a bar plot for each site. 
a. What are the three most common sites to observe cancer after 8 transitions? 

```{r}
# code goes here


```

*Comment goes here*

### Excerise 5:

a. Compare this to the exact probability using direct matrix multiplication. 
a. Comment of the difference between the simulation and exact probability. 

```{r}
library(expm)


```

*Comment goes here*

### Excerise 6:

a. Extend this above to 1000 MC draws from the Markov chain simulation to calculate the mean first passage. 
a. Generate a summary table of the results displaying the states with the 5 shortest mean first passage times.
a. What insight might you be able to give medical research. 

```{r}

```

*Comment goes here*
